```{r libraries}
library(tidyverse)
library(dplyr)
library(readr)
library(janitor)
```

```{r loading-data}
housing_raw <- read_csv("data/Melbourne_housing_FULL.csv")
problems(read_csv("data/Melbourne_housing_FULL.csv"))
```

```{r cleaning-data}
housing<- read_csv("data/Melbourne_housing_FULL.csv",
                        na = c("", "NA", "#N/A")) |>
  clean_names() |>
  distinct() |>
  # response variable must be present so we can drop 3 rows with no region
  # without about 22% are dropped.
  # rows with missing region/council info were removed
  filter(!is.na(price), !is.na(regionname)) |>
  # fix misspelled source column names and use clearer names
  rename(latitude = lattitude, longitude = longtitude,
         region = regionname, bedrooms = bedroom2) |>
  mutate(
     # dates are stored as day/month/year 
    date = dmy(date),
    sale_year = factor(year(date)),
    # replace single-letter codes with readable labels
    type = factor(type, levels = c("h", "u", "t"),
                  labels = c("House", "Unit", "Townhouse")),
    method = factor(method,
                    levels = c("S", "SP", "PI", "VB", "SA"),
                    labels = c("Sold", "Sold prior", "Passed in",
                               "Vendor bid", "Sold after auction")),
    # the three "Victoria" regions together have only ~430 sales, so they
    # are combined into one "Regional Victoria" level for filtering reasons
    region_group = if_else(str_detect(region, "Victoria"),
                           "Regional Victoria", region) |> factor(),
    region = factor(region),
    # placeholder values -> NA
    # land size of 0 is a placeholder, not a real value
    landsize      = na_if(landsize, 0),
    # for some units, landsize is the land under the whole building/complex
    # (some 2-room units show 15,000+ m^2),so not meaningful
    landsize = if_else(type == "Unit", NA_real_, landsize),
    # building areas of 0 or > 5,000 m^2 are data-entry errors
    building_area = if_else(building_area == 0 | building_area > 5000,
                            NA_real_, building_area),
    # year built must fall between 1800 and the end of the data (2018)
    # values like 1196 and 2106 are typos
    year_built    = if_else(year_built < 1800 | year_built > 2018,
                            NA_real_, year_built),
    # postcode is an identifier
    postcode = as.character(postcode)
  ) |>

# Keep derived columns next to the variables they come from
relocate(sale_year, .after  = date) |>
relocate(region_group, .after = region)
```

```{r sanity-check}
# look into the clean dataset
glimpse(housing)
# check for NA values in the columns
colSums(is.na(housing))

# two-way table preview
tabyl(housing, type, region_group)          
housing |> group_by(type) |>
  summarize(across(c(price, landsize, building_area),
                   list(mean = ~mean(.x, na.rm = TRUE),
                        median = ~median(.x, na.rm = TRUE))))
```

## Contingency Tables

### One-way tables
```{r one-way-tidy}
# tidyverse: counts of each property type
housing |>
  drop_na(type) |>
  group_by(type) |>
  summarize(count = n())

# counts of each sale method
housing |>
  drop_na(method) |>
  group_by(method) |>
  summarize(count = n())
```


```{r one-way-tabyl}
# tabyl: same counts, with proportions included automatically
housing |> tabyl(type)
housing |> tabyl(method) |> adorn_pct_formatting()
```

Houses make up about two-thirds of the sales (18,470 of 27,244), and most
properties sold outright ("Sold", 64%).

## Two-way tables

```{r two-way-tidy}
# tidyverse: region group by property type, pivoted to wide format
housing |>
  drop_na(region_group, type) |>
  group_by(region_group, type) |>
  summarize(count = n(), .groups = "drop") |>
  pivot_wider(names_from = type, values_from = count)
```


```{r two-way-tabyl}
# tabyl: same table, plus row/column totals
housing |>
  tabyl(region_group, type) |>
  adorn_totals(c("row", "col"))

# Row percentages: the mix of property types within each region
housing |>
  tabyl(region_group, type) |>
  adorn_percentages("row") |>
  adorn_pct_formatting()
```

Southern Metropolitan has the highest share of units (33%), while Regional
Victoria is almost entirely houses (97%).

### Three-way table

```{r three-way-tidy}
# tidyverse: region group by type by sale year (year across the columns)
housing |>
  drop_na(region_group, type, sale_year) |>
  group_by(region_group, type, sale_year) |>
  summarize(count = n(), .groups = "drop") |>
  pivot_wider(names_from = sale_year, values_from = count)
```


```{r three-way-tabyl}
# tabyl: returns a list with one two-way table for each sale year
housing |> 
  tabyl(region_group, type, sale_year)
```

## Numerical Summaries

### Price at each level of property type


```{r price-by-type}
# center and spread of price by property type
housing |>
  group_by(type) |>
  summarize(n = n(),
            mean_price = mean(price, na.rm = TRUE),
            median_price = median(price, na.rm = TRUE),
            sd_price = sd(price, na.rm = TRUE),
            min_price = min(price, na.rm = TRUE),
            max_price = max(price, na.rm = TRUE)
  )
```

### Price at each level of region group

```{r price-by-region}
# center and spread of price by region group
housing |>
  group_by(region_group) |>
  summarize(n = n(),
            mean_price = mean(price),
            median_price = median(price),
            sd_price = sd(price)) |>
  arrange(median_price)
```

### Median price by region and property type


```{r price-by-region-type}
# median price by region/type combination, pivoted to wide formart
housing |>
  group_by(region_group, type) |>
  summarize(median_price = median(price), .groups = "drop") |>
  pivot_wider(names_from = type, values_from = median_price)
```


```{r other-variables-by-type}
# land size is NaN for units because it was set to missing during cleaning
# many variables have NAs so na.rm = TRUE is used to avoid errors in the summary
housing |>
  group_by(type) |>
  summarize(across(c(rooms, distance, landsize, building_area, year_built),
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        median = ~ median(.x, na.rm = TRUE),
                        sd = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.col}_{.fn}"))
```

## Plots


```{r plot-setup}
theme_set(theme_minimal()) # sets all plots with the same minimal theme
```

### 1. Bar chart: sale method

```{r plot-method-bar}
housing |>
  ggplot(aes(x = fct_infreq(method))) +
  geom_bar(fill = "steelblue") + 
  labs(title = "Number of Sales by Sale Method", 
  x = "Sales Method", y = "Number of Sales") 
```

Most properties (64%) sold, while selling after auction is rare (under 1%).


### 2. Side-by-side bar chart: property type by region group


```{r plot-region-type-bar}
# count of each property type within each region group
housing |>
  ggplot(aes(y = region_group, fill = type)) +
  geom_bar(position = "dodge") +
  labs(title = "Property Type by Region", x = "Number of sales", y = "Region", fill = "Property Type")
```

Northern Metropolitan has the most house sales (5,309), but Southern Metropolitan has the most overall (8,524) since of its many townhouses. The share of units is highest in Southern Metropolitan (33%) and lowest in Regional Victoria (2%).

### 3. Faceted histogram: price by property type

```{r plot-price-histogram}
housing |>
  ggplot(aes(x = price, fill = type)) +
  geom_histogram(bins = 50, show.legend = FALSE) +
  facet_wrap(~ type, ncol = 1, scales = "free_y") +
  coord_cartesian(xlim = c(0,4e6)) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Distribution of Sale Price by Property Type", x = "Sale Price (AUD)", y = "Number of sales")
```

All three distributions are right-skewed. Units cluster at lower prices, while house prices are much more spread out.


### 4. Boxplot: price by property type

```{r plot-price-boxplot}
# Log scale makes the skewed prices easier to compare across groups
housing |>
  ggplot(aes(x = price, y = region_group, fill = type)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_x_log10(labels = scales::dollar_format())+
  labs(title = "Sale Price by Region and Property Type",
       x = "Sale price (AUD, log scale)", y = "Region", fill = "Property type")
```

In the metropolitan regions, houses have the highest median price and units the lowest, with the largest gap in Southern Metropolitan. Regional Victoria has very few townhouse (3) and unit (10) sales, so its boxes are not reliable comparisons.

### 5. Scatter plot: price vs. distance from the CBD

```{r plot-price-distance-scatter}
# Transparency helps with ~27,000 overlapping points
housing |>
  ggplot(aes(x = distance, y = price, color = type)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_log10(labels = scales::dollar_format()) +
  labs(title = "Sale Price vs. Distance from Melbourne CBD",
       x = "Distance from CBD (km)", y = "Sale price (AUD, log scale)",
       color = "Property type")
```

House prices fall noticeably with distance from the city (about 3% per km), townhouse prices fall slightly, and unit prices show almost no relationship with distance.

### 6. Heatmap: median price by region and number of rooms

```{r plot-price-heatmap}
# Group 5+ rooms together since few properties have more than 5.
# Each cell shows the median price and the number of sales behind it.
housing |>
  mutate(rooms_group = if_else(rooms >= 5, "5+", as.character(rooms)),
         rooms_group = factor(rooms_group, levels = c("1", "2", "3", "4", "5+"))) |>
  group_by(region_group, rooms_group) |>
  summarize(median_price = median(price), n = n(), .groups = "drop") |>
  ggplot(aes(x = rooms_group, y = region_group, fill = median_price)) +
  geom_tile(color = "white") +
    geom_text(aes(label = paste0(scales::label_dollar(scale = 1e-3, suffix = "K",
                                                    accuracy = 1)(median_price),
                               "\n(n = ", n, ")"),
                color = if_else(median_price > 1.5e6, "black", "white")),
            size = 2.8) +
  scale_color_identity() +
  scale_fill_viridis_c(labels = scales::dollar_format()) +
  labs(title = "Median Sale Price by Region and Number of Rooms",
       x = "Number of rooms", y = "Region", fill = "Median price")
```

In the metropolitan regions, median price rises steadily with the number of rooms, most steeply in Southern Metropolitan ($405K for 1 room to $2.4M for 5+). The high Regional Victoria value for 1 room is based on only 2 sales.