```{r libraries}
library(tidyverse)
library(dplyr)
library(readr)
library(janitor)
```

```{r loading-data}
housing_raw <- read_csv("data/Melbourne_housing_FULL.csv")
problems(read_csv("data/Melbourne_housing_FULL.csv"))
```

```{r cleaning-data}
housing<- read_csv("data/Melbourne_housing_FULL.csv",
                        na = c("", "NA", "#N/A")) |>
  clean_names() |>
  distinct() |>
  # response variable must be present so we can drop 3 rows with no region
  # without about 22% are dropped.
  # rows with missing region/council info were removed
  filter(!is.na(price), !is.na(regionname)) |>
  # fix misspelled source column names and use clearer names
  rename(latitude = lattitude, longitude = longtitude,
         region = regionname, bedrooms = bedroom2) |>
  mutate(
     # dates are stored as day/month/year 
    date = dmy(date),
    sale_year = factor(year(date)),
    # replace single-letter codes with readable labels
    type = factor(type, levels = c("h", "u", "t"),
                  labels = c("House", "Unit", "Townhouse")),
    method = factor(method,
                    levels = c("S", "SP", "PI", "VB", "SA"),
                    labels = c("Sold", "Sold prior", "Passed in",
                               "Vendor bid", "Sold after auction")),
    # the three "Victoria" regions together have only ~430 sales, so they
    # are combined into one "Regional Victoria" level for filtering reasons
    region_group = if_else(str_detect(region, "Victoria"),
                           "Regional Victoria", region) |> factor(),
    region = factor(region),
    # placeholder values -> NA
    # land size of 0 is a placeholder, not a real value
    landsize      = na_if(landsize, 0),
    # for some units, landsize is the land under the whole building/complex
    # (some 2-room units show 15,000+ m^2),so not meaningful
    landsize = if_else(type == "Unit", NA_real_, landsize),
    # building areas of 0 or > 5,000 m^2 are data-entry errors
    building_area = if_else(building_area == 0 | building_area > 5000,
                            NA_real_, building_area),
    # year built must fall between 1800 and the end of the data (2018)
    # values like 1196 and 2106 are typos
    year_built    = if_else(year_built < 1800 | year_built > 2018,
                            NA_real_, year_built),
    # postcode is an identifier
    postcode = as.character(postcode)
  ) |>

# Keep derived columns next to the variables they come from
relocate(sale_year, .after  = date) |>
relocate(region_group, .after = region)
```

```{r sanity-check}
# look into the clean dataset
glimpse(housing)
# check for NA values in the columns
colSums(is.na(housing))

# two-way table preview
tabyl(housing, type, region_group)          
housing |> group_by(type) |>
  summarize(across(c(price, landsize, building_area),
                   list(mean = ~mean(.x, na.rm = TRUE),
                        median = ~median(.x, na.rm = TRUE))))
```

## Contingency Tables

### One-way tables
```{r one-way-tidy}
# tidyverse: counts of each property type
housing |>
  drop_na(type) |>
  group_by(type) |>
  summarize(count = n())

# counts of each sale method
housing |>
  drop_na(method) |>
  group_by(method) |>
  summarize(count = n())
```


```{r one-way-tabyl}
# tabyl: same counts, with proportions included automatically
housing |> tabyl(type)
housing |> tabyl(method) |> adorn_pct_formatting()
```

Houses make up about two-thirds of the sales (18,470 of 27,244), and most
properties sold outright ("Sold", 64%).

## Two-way tables

```{r two-way-tidy}
# tidyverse: region group by property type, pivoted to wide format
housing |>
  drop_na(region_group, type) |>
  group_by(region_group, type) |>
  summarize(count = n(), .groups = "drop") |>
  pivot_wider(names_from = type, values_from = count)
```


```{r two-way-tabyl}
# tabyl: same table, plus row/column totals
housing |>
  tabyl(region_group, type) |>
  adorn_totals(c("row", "col"))

# Row percentages: the mix of property types within each region
housing |>
  tabyl(region_group, type) |>
  adorn_percentages("row") |>
  adorn_pct_formatting()
```

Southern Metropolitan has the highest share of units (33%), while Regional
Victoria is almost entirely houses (97%).

### Three-way table

```{r three-way-tidy}
# tidyverse: region group by type by sale year (year across the columns)
housing |>
  drop_na(region_group, type, sale_year) |>
  group_by(region_group, type, sale_year) |>
  summarize(count = n(), .groups = "drop") |>
  pivot_wider(names_from = sale_year, values_from = count)
```


```{r three-way-tabyl}
# tabyl: returns a list with one two-way table for each sale year
housing |> 
  tabyl(region_group, type, sale_year)
```

## Numerical Summaries

### Price at each level of property type


```{r price-by-type}
# center and spread of price by property type
housing |>
  group_by(type) |>
  summarize(n = n(),
            mean_price = mean(price, na.rm = TRUE),
            median_price = median(price, na.rm = TRUE),
            sd_price = sd(price, na.rm = TRUE),
            min_price = min(price, na.rm = TRUE),
            max_price = max(price, na.rm = TRUE)
  )
```

### Price at each level of region group

```{r price-by-region}
# center and spread of price by region group
housing |>
  group_by(region_group) |>
  summarize(n = n(),
            mean_price = mean(price),
            median_price = median(price),
            sd_price = sd(price)) |>
  arrange(median_price)
```

### Median price by region and property type


```{r price-by-region-type}
# median price by region/type combination, pivoted to wide formart
housing |>
  group_by(region_group, type) |>
  summarize(median_price = median(price), .groups = "drop") |>
  pivot_wider(names_from = type, values_from = median_price)
```


```{r}
# land size is NaN for units because it was set to missing during cleaning
# many variables have NAs so na.rm = TRUE is used to avoid errors in the summary
housing |>
  group_by(type) |>
  summarize(across(c(rooms, distance, landsize, building_area, year_built),
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        median = ~ median(.x, na.rm = TRUE),
                        sd = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.col}_{.fn}"))
```